Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)

### 2020-07-09 manuscript/manuscript.Rmd
manuscript v0.6

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: elsevier-vancouver.csl
output:
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
  word_document:
    toc: yes
    reference_docx: template.docx
link-citations: yes
always_allow_html: yes
---

@@

reference_homes <- reference_homes %>% 
  filter(home_code %in% home_inclusion) %>% 
  left_join(select(filter(beds, year=="2020"), -year)) %>% 
  mutate(staff_to_bed_ratio = staff_total/beds,
         carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
         bed_tot_category = cut(.$beds, 
           breaks = seq(20,100,25), right = F, 
           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
         )

@@

residents_overview <- tline %>% 
  filter(rday==1) %>%
  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  group_by(resident_id, home_code, ageg, gender, 
           admission_type2, home_type, user_type2, status) %>% 
  summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed),
            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
            confirmed = max(-susceptible_confirmed+1, na.rm = T),
            start_date = min(date, na.rm = T),
            stop_date = max(date, na.rm = T),
            start_int = as.numeric(min(date, na.rm = T) - date_min),
            stop_int = as.numeric(max(date, na.rm = T)- date_min)
    ) %>% 
  ungroup() %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category,
                   beds,  pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))

@@

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  ))  %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date+1),
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric( entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed")),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, carer_to_bed_ratio, bed_tot_category, 
                   beds, pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% 
  ungroup()

@@

residents_episodes_short <- residents_episodes %>% 
  filter(exit_date > date_min_cox_mortality) %>% 
  mutate(date1 = if_else(
    entry_date > date_min_cox_mortality,
    date1,
    as.numeric( date_min_cox_mortality - date_min),
  ))

@@

incidence_CT_residents <- table_nums(
  name = "incidence_CT_residents",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
supp_epi_CT_oubreak_only <- table_nums(
  name = "incidence_CT_staff",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_resident_descriptives <- table_nums(
  name = "tab_resident_descriptives",
  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CFR_age_sex <- table_nums(
  name = "tab_CFR_age_sex",
  caption =  paste0("All-cause case-fatality rates and relative risks by age and sex among private residents (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  "; ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CoxPH_confirmed <- table_nums(
  name = "tab_CoxPH_confirmed",
  caption =  paste0("Risk factors for confirmed infection in private residents: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  ")"))
tab_CoxPH_death <- table_nums(
  name = "tab_CoxPH_death",
  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_attributable_deaths <- table_nums(
  name = "tab_attributable_deaths",
  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
  formatbm(dplyr::n_distinct(residents_episodes_short$resident_id)),
  ", ", format(date_min_cox_mortality, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))

@@

```{r table_resident_descriptives, message = FALSE, results="asis"}
residents_overview$outbreak <- relevel(as.factor(dplyr::case_when(
  residents_overview$home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
  TRUE ~ "Other homes"
)), "Outbreak homes")

arsenal::tableby(
  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Home type` +  
    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
  data = mutate(residents_overview,
                Sex = gender,
                Age = ageg,
                `Resident type` = user_type2,
                `Admission type` = admission_type2,
                `Home type` = home_type,
                `Infection status by 14 June` = covid_status, 
                `Status as of 14 June` = status,
                `Region/nation` = region_UK 
  )) %>% summary()
```

@@

```{r tab_CFR_age_sex, results="asis"}
residents_overview %>%
  mutate(Age = ageg, Sex = gender) %>% 
  group_by(Sex, Age) %>% 
  summarise(
    n = n_distinct(resident_id),
    resident_days = sum(resident_days),
    confirmed_infections = sum(confirmed == 1),
    deaths = sum(status == "Deceased"),
    deaths_confirmed = sum(status == "Deceased" & confirmed == 1)) %>% 
  mutate(
    cfr = tbrounding(deaths_confirmed/confirmed_infections*100),
    cfr_lower = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$lower*100),
    cfr_upper = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$upper*100),
    RR = relative_risk(deaths_confirmed, confirmed_infections - deaths_confirmed,
                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed)),
    RR_SE = relative_risk_SE(deaths_confirmed, confirmed_infections - deaths_confirmed,
                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed))
  ) %>%  ungroup() %>% 
  transmute(
    Age,
    Sex,
    N = formatbm(n),
    `Confirmed infections` = confirmed_infections,
    `Total deaths` = deaths,
    `Deaths in confirmed infections` = deaths_confirmed,
    `Case-fatality rate (%)` = paste0(cfr, " [", cfr_lower, "; ", cfr_upper, "]")
    # `Relative risk` = paste0(
    #   tbrounding(RR, 2), " [",
    #   tbrounding(exp(log(RR) - 1.96 * RR_SE), 2), "; ",
    #   tbrounding(exp(log(RR) + 1.96 * RR_SE), 2), "]"
    # )
  ) %>% kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
```

@@

Higher staff to resident ratios were associated with lower risk of infection: a 10 percentage point increase in the number of staff relative to the total number of care home beds was associated with a 23% reduction [18%, 29%] in the hazards of confirmed infection, once adjusting for other factors.

@@

```{r hazard_ratios_confirmed,   results="asis"}
# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
#  summary(cph_conf)
# cox.zph(cph_conf)
# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")

explanatory = c("gender", "ageg", "home_type", "user_type2",  
                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6"  )
dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
CPH_confirmed <-  finalfit(residents_overview,
                           dependent, 
                           explanatory, column = T,
                           add_dependent_label = F)  
knitr::kable(CPH_confirmed,
             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

```{r residents_daily}
residents_daily <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  )) %>% 
  group_by(resident_id) %>% 
  mutate(death = case_when(
    status == "Deceased" & date == max(date) ~ 1L,
    TRUE ~ 0L
  )) %>% 
  ungroup() %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1)) %>% 
  mutate(covid_status_outbreak = as.factor(case_when(
    symptomatic==0 & confirmed == 0 & outbreak == 0 ~ "0 Uninfected (other home)",
    symptomatic==1 & confirmed == 0 & outbreak == 0 ~ "1 Symptomatic not confirmed  (other home)",
    symptomatic==0 & confirmed == 0 & outbreak == 1 ~ "2 Uninfected (outbreak home)",
    symptomatic==1 & confirmed == 0 & outbreak == 1 ~ "3 Symptomatic not confirmed  (outbreak home)",
    symptomatic==0 & confirmed == 1 ~ "4 Confirmed asymptomatic",
    symptomatic==1 & confirmed == 1 ~ "5 Confirmed symptomatic"
  )))
residents_daily$covid_status_outbreak_digit <- as.factor(
  as.numeric(residents_daily$covid_status_outbreak)-1)
residents_daily <- bind_cols(
  residents_daily,
  as_tibble(model.matrix(~covid_status_outbreak_digit, residents_daily)[,-1])
)

@@

```{r hazard_ratios_death, results="asis"}
# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
#                       gender+ ageg + home_type + covid_status_outbreak +
#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
#                       cluster(resident_id), data = residents_episodes_short, ties = "breslow")
# summary(cph_death2)
# cox.zph(cph_death2)

explanatory = c("gender", "ageg", "home_type", "user_type2",
                "staff_to_bed_ratio", "bed_tot_category", "occ_bedroom_ratio_g6",
                "covid_status_outbreak",  "cluster(resident_id)")
dependent = "Surv(time = date1, time2 = date2, event = death_status)"
CPH_death <-  finalfit(residents_episodes_short, dependent, explanatory, 
                       column = T, add_dependent_label = F)  
knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

```{r fig_KM_mortality,  fig.height = 6, fig.width = 8, message=F}
KM_mortality_by_covid_status <- survfit(
  Surv(time = date1, time2 = date2, event = death_status) ~ covid_status + outbreak, 
  data = residents_episodes_short)
KM_mortality_plot <- ggsurvplot_facet(KM_mortality_by_covid_status,
                 residents_episodes_short,
                 facet.by  = "covid_status",
                 legend.labs = c("Non-outbreak home", "Outbreak home"),
                 short.panel.labs	 = T, break.x.by	= 7, 
                 break.y.by = .1,
                 palette = "aaas", conf.int = T,
                 scales = "free", ggtheme = theme_bw(), xlim = c(32, 104),
                 xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
                 conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)
pdf("figure5.pdf",8, 6)
KM_mortality_plot
dev.off()
KM_mortality_plot
```

*Note*: The SARS-COV-2 classification is time-dependent and applies from the date of incident recorded on the relevant Datix incident report.

@@

```{r attributable_deaths, warning=F, message=F, results = "asis"}
cph_death3 <- coxph(Surv(date1, date2, death_status) ~
                    gender+ ageg + home_type + covid_status_outbreak_digit +
                    staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
                    cluster(resident_id),
                    data =  residents_episodes_short)

@@

attributable_deaths <- residents_episodes_short %>%
  group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
  summarise(
    rdays = sum(as.integer(date2-date1)),
    deaths = sum(death_status == 1)
    ) %>% 
  left_join(HRs_covid) %>% 
  mutate(HR = tidyr::replace_na(HR, 1),
           HR_lower = tidyr::replace_na(HR_lower, 1),
           HR_upper = tidyr::replace_na(HR_upper, 1))

death_predictions <- 1-predict(cph_death3, residents_episodes_short, type = "survival")

>>>>

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: elsevier-vancouver.csl
output:
  word_document:
    toc: yes
    reference_docx: template.docx
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
link-citations: yes
always_allow_html: yes
---

@@

reference_homes <- reference_homes %>% 
  filter(home_code %in% home_inclusion) %>% 
  left_join(select(filter(beds, year=="2020"), -year)) %>% 
  mutate(staff_to_bed_ratio = staff_total/beds,
         carer_to_bed_ratio = (staff_care_assistants + staff_nurses)/beds,
         bed_tot_category = cut(.$beds, 
           breaks = seq(20,100,25), right = F, 
           paste0(seq(20, 80, 25), "–", seq(34, 94, 25), " beds"))
         ) %>% 
  mutate(staff_to_bed_ratio_g5 = cut(
    staff_to_bed_ratio, c(min(staff_to_bed_ratio)-.001,
    quantile(staff_to_bed_ratio, seq(.2, .8, .2)),
    max(staff_to_bed_ratio))))

@@

residents_overview <- tline %>% 
  filter(rday==1) %>%
  mutate(ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  group_by(resident_id, home_code, ageg, gender, 
           admission_type2, home_type, user_type2, status) %>% 
  summarise(resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed),
            symptomatic = max(-susceptible_symptomatic+1, na.rm = T),
            confirmed = max(-susceptible_confirmed+1, na.rm = T),
            start_date = min(date, na.rm = T),
            stop_date = max(date, na.rm = T),
            start_int = as.numeric(min(date, na.rm = T) - date_min),
            stop_int = as.numeric(max(date, na.rm = T)- date_min)
    ) %>% 
  ungroup() %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed"))) %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category,
                   beds,  pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  left_join(distinct(select(incidents, resident_id, starts_with("covid_first"))))

@@

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  ))  %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date+1),
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric( entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(covid_status = factor(covid_status, 
                               levels = c("Uninfected", "Symptomatic (not confirmed)",
                                          "Asymptomatic confirmed", "Symptomatic confirmed")),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95))   %>% 
  left_join(select(reference_homes, home_code, country, region_UK,
                   staff_to_bed_ratio, staff_to_bed_ratio_g5, carer_to_bed_ratio, bed_tot_category, 
                   beds, pct_occupancy, no_bedrooms, 
                   occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% 
  ungroup()

@@

residents_episodes_short <- residents_episodes %>% 
  filter(exit_date > date_min_cox_mortality) %>% 
  mutate(date1 = if_else(
    entry_date >= date_min_cox_mortality,
    date1,
    as.numeric( date_min_cox_mortality - date_min),
  ))

@@

incidence_CT_residents <- table_nums(
  name = "incidence_CT_residents",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
supp_epi_CT_oubreak_only <- table_nums(
  name = "incidence_CT_staff",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_resident_descriptives <- table_nums(
  name = "tab_resident_descriptives",
  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CFR_age_sex <- table_nums(
  name = "tab_CFR_age_sex",
  caption =  paste0("All-cause case-fatality rates by age and sex among private residents (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  "; ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CoxPH_confirmed <- table_nums(
  name = "tab_CoxPH_confirmed",
  caption =  paste0("Risk factors for confirmed infection in private residents: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  ")"))
tab_CoxPH_death <- table_nums(
  name = "tab_CoxPH_death",
  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_attributable_deaths <- table_nums(
  name = "tab_attributable_deaths",
  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))

@@

```{r table_resident_descriptives, message = FALSE, results="asis"}

residents_overview <- residents_overview %>% 
  mutate(outbreak = as.integer(home_code %in% desc_residents_outbreak_only$outbreak_homes) ) %>% 
  dplyr::mutate(
    covid_status_outbreak = as.factor(case_when(
    covid_status == "Uninfected" & outbreak == 0 ~ "0 Uninfected (other home)",
    covid_status == "Symptomatic (not confirmed)" &
      outbreak==0 ~ "1 Symptomatic not confirmed  (other home)",
    covid_status == "Uninfected" & outbreak == 1 ~ "2 Uninfected (outbreak home)",
    covid_status == "Symptomatic (not confirmed)" &
      outbreak==1 ~ "3 Symptomatic not confirmed  (outbreak home)",
    covid_status == "Asymptomatic confirmed" ~ "4 Confirmed asymptomatic",
    covid_status == "Symptomatic confirmed" ~ "5 Confirmed symptomatic"
  ))) %>% 
  mutate(outbreak = relevel(as.factor(dplyr::case_when(
    home_code %in% desc_residents_outbreak_only$outbreak_homes ~ "Outbreak homes",
    TRUE ~ "Other homes"
    )), "Outbreak homes"))

arsenal::tableby(
  outbreak  ~ Sex + Age  +  `Resident type` + `Admission type`  + `Funding type` +  
    `Infection status by 14 June` + `Status as of 14 June` + `Region/nation`, 
  control = arsenal::tableby.control(test = FALSE, numeric.stats = c("Nmiss", "meansd"), digits = 1L),
  data = mutate(residents_overview,
                Sex = gender,
                Age = ageg,
                `Resident type` = user_type2,
                `Admission type` = admission_type2,
                `Funding type` = home_type,
                `Infection status by 14 June` = covid_status, 
                `Status as of 14 June` = status,
                `Region/nation` = region_UK 
  )) %>% summary()
```

@@

```{r tab_CFR_age_sex, results="asis"}
CFR_age_sex <- residents_overview %>%
  mutate(Age = ageg, Sex = gender) %>% 
  group_by(Sex, Age) %>% 
  summarise(
    n = n_distinct(resident_id),
    resident_days = sum(resident_days),
    confirmed_infections = sum(confirmed == 1),
    deaths = sum(status == "Deceased"),
    deaths_confirmed = sum(status == "Deceased" & confirmed == 1)) %>% 
  ungroup()

CFR_age_sex <- bind_rows(
  CFR_age_sex,
  colSums(CFR_age_sex[3:7])
) %>% mutate(
  Sex = tidyr::replace_na(Sex, "All"),
  Age = tidyr::replace_na(Sex, "")
)

CFR_age_sex <- CFR_age_sex %>% 
  mutate(
    cfr = tbrounding(deaths_confirmed/confirmed_infections*100),
    cfr_lower = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$lower*100),
    cfr_upper = tbrounding(epitools::binom.exact(deaths_confirmed, confirmed_infections)$upper*100),
    RR = relative_risk(deaths_confirmed, confirmed_infections - deaths_confirmed,
                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed)),
    RR_SE = relative_risk_SE(deaths_confirmed, confirmed_infections - deaths_confirmed,
                       deaths - deaths_confirmed, n - confirmed_infections -(deaths - deaths_confirmed))
  ) %>%  ungroup() %>% 
  transmute(
    Age,
    Sex,
    N = formatbm(n),
    `Confirmed infections` = confirmed_infections,
    `Total deaths` = formatbm(deaths),
    `Deaths in confirmed infections` = deaths_confirmed,
    `Case-fatality rate (%)` = paste0(cfr, " [", cfr_lower, "; ", cfr_upper, "]")
    # `Relative risk` = paste0(
    #   tbrounding(RR, 2), " [",
    #   tbrounding(exp(log(RR) - 1.96 * RR_SE), 2), "; ",
    #   tbrounding(exp(log(RR) + 1.96 * RR_SE), 2), "]"
    # )
  ) 
CFR_age_sex %>%
  kable(align=c("l", "l", "r", "r", "r", "r", "r", "r"))
```

@@

Higher staff to resident ratios were associated with lower risk of infection: care homes with less than 1 staff per bed had a risk of infection four times higher than those with >=1.4 staff per bed, once adjusting for other factors.

@@

```{r hazard_ratios_confirmed,   results="asis"}
# cph_conf <- coxph(Surv(time = start_int, time2 = stop_int +1, event = confirmed) ~ gender + ageg+home_type+user_type2+  staff_to_bed_ratio_g5+bed_tot_category  +occ_bedroom_ratio_g6, data = residents_overview)
#  summary(cph_conf)
# cox.zph(cph_conf)
# survfit(Surv(time = start_int, time2 = stop_int +1, event = confirmed)~factor(gender), data = residents_overview) %>% ggsurvplot(ylim=c(.85,1), conf.int = T, legend = "right")

explanatory = c("gender", "ageg", "home_type", "user_type2",  
                "bed_tot_category", "staff_to_bed_ratio_g5",  "occ_bedroom_ratio_g6")
dependent = "Surv(time = start_int, time2 = stop_int +1, event = confirmed)"
CPH_confirmed <-  finalfit(residents_overview,
                           dependent, 
                           explanatory, column = T,
                           add_dependent_label = F)  
knitr::kable(CPH_confirmed,
             row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

```{r residents_daily}
residents_daily <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_first_case) %>%
  mutate(outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  )) %>% 
  group_by(resident_id) %>% 
  mutate(
    death = case_when(
    status == "Deceased" & date == max(date) ~ 1L,
    TRUE ~ 0L
  ),death_Apr = case_when(
    status == "Deceased" & date == max(date) & date >= as.Date("2020-04-01") ~ 1L,
    TRUE ~ 0L
  )) %>% 
  ungroup() %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1)) %>% 
  mutate(covid_status_outbreak = as.factor(case_when(
    symptomatic==0 & confirmed == 0 & outbreak == 0 ~ "0 Uninfected (other home)",
    symptomatic==1 & confirmed == 0 & outbreak == 0 ~ "1 Symptomatic not confirmed  (other home)",
    symptomatic==0 & confirmed == 0 & outbreak == 1 ~ "2 Uninfected (outbreak home)",
    symptomatic==1 & confirmed == 0 & outbreak == 1 ~ "3 Symptomatic not confirmed  (outbreak home)",
    symptomatic==0 & confirmed == 1 ~ "4 Confirmed asymptomatic",
    symptomatic==1 & confirmed == 1 ~ "5 Confirmed symptomatic"
  )))
residents_daily$covid_status_outbreak_digit <- as.factor(
  as.numeric(residents_daily$covid_status_outbreak)-1)
residents_daily <- bind_cols(
  residents_daily,
  as_tibble(model.matrix(~covid_status_outbreak_digit, residents_daily)[,-1])
)

@@

```{r hazard_ratios_death, results="asis"}
# cph_death2 <- coxph(Surv(date1, date2, death_status) ~
#                       gender+ ageg + home_type + covid_status_outbreak +
#                       staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
#                       cluster(resident_id), data = residents_episodes)
# summary(cph_death2)
# cox.zph(cph_death2)

explanatory = c("gender", "ageg", "home_type", "user_type2",
                "bed_tot_category", "staff_to_bed_ratio_g5", "occ_bedroom_ratio_g6",
                "covid_status_outbreak",  "cluster(resident_id)")
dependent = "Surv(time = date1, time2 = date2, event = death_status)"
CPH_death <-  finalfit(residents_episodes, dependent, explanatory, 
                       column = T, add_dependent_label = F)  
knitr::kable(CPH_death, row.names=FALSE, align=c("l", "l", "r", "r", "r", "r"))
```

@@

```{r fig_KM_mortality,  fig.height = 6, fig.width = 8, message=F}
KM_mortality_by_covid_status <- survfit(
  Surv(time = start_int, time2 = stop_int, event = death_status) ~ covid_status + outbreak, 
  data = mutate(
    residents_overview, 
    death_status = as.integer(status == "Deceased")
  ))
KM_mortality_plot <- ggsurvplot_facet(
  KM_mortality_by_covid_status,
  mutate(residents_overview, 
         death_status = as.integer(status == "Deceased")  ),
  facet.by  = "covid_status",
  legend.labs = c("Non-outbreak home", "Outbreak home"),
  short.panel.labs	 = T, break.x.by	= 7, 
  break.y.by = .1,
  palette = "aaas", conf.int = T,
  scales = "free", ggtheme = theme_bw(), xlim = c(0, 104), ylim = c(.4, 1),
  xlab  = paste0("Days since ", format(date_min, "%e %b %Y")),
  conf.int.alpha = .15, censor.shape = 124, censor.size = 2, size = 0.5)

KM_mortality_plot <- KM_mortality_plot +  
  scale_x_continuous(name = "Date", 
                     breaks = c(seq(0, 7*50, 7)),
                     labels = format.Date(date_min + seq(0, 7*50, 7), "%e %b %Y" )) +
  theme(axis.text.x = element_text(size = 8, angle = 35, vjust = 1, hjust=1)) 
pdf("figure3.pdf", 8, 6)
KM_mortality_plot
dev.off()
KM_mortality_plot
```

@@

```{r attributable_deaths, warning=F, message=F, results = "asis"}
cph_death3 <- coxph(Surv(date1, date2, death_status) ~
                    gender+ ageg + home_type + covid_status_outbreak_digit +
                    staff_to_bed_ratio + bed_tot_category + occ_bedroom_ratio_g6 +
                    cluster(resident_id), data =  residents_episodes)

@@

attributable_deaths <- residents_episodes %>%
  group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
  summarise(
    rdays = sum(as.integer(date2-date1)),
    deaths = sum(death_status == 1),
    deaths_apr = sum(death_status == 1 & date2 >= 30)
    ) %>% 
  left_join(HRs_covid) %>% 
  mutate(HR = tidyr::replace_na(HR, 1),
           HR_lower = tidyr::replace_na(HR_lower, 1),
           HR_upper = tidyr::replace_na(HR_upper, 1))

# All deaths
